---
title: "alcohol-related-admissions"
author: "D Blana"
date: "17/09/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load packages

We just need the tidyverse package.

```{r load-packages}
library(tidyverse)
```

## Read in data
Our data is saved in csv files, so we will use the read_csv function.

```{r read-data}
# alcohol-related hospital admissions
data <- read_csv("scotpho_data_extract.csv") 

# intermediate data zone info
intermediate_zone_codes <- read_csv("iz2011_codes_and_labels_21042020.csv") 

```


## Prepare data
The data in these open datasets is already clean, so we don't need to worry about missing or wrong values. It is also tidy. 
We will only select the data we want: data from intermediate zones, and a subset of the variables from each dataset.

```{r}
glimpse(data)

iz_data <- data %>%
  filter(area_type == "Intermediate zone") %>%
  select(area_name,area_code,year,measure)

glimpse(intermediate_zone_codes)

iz_info <- intermediate_zone_codes %>%
  select(IntZone,HBName)

```

## Join the datasets
Let's join the two datasets, so we know which health board each data zone belongs to.
We will also remove "NHS" from the health board name, and rename a couple of the columns.

```{r join-datasets}
admission_data <- left_join(iz_data,iz_info, by = c("area_code" = "IntZone"))

glimpse(admission_data)

admission_data <- admission_data %>%
  mutate(HBName = gsub("NHS ", "", HBName)) %>%
  rename(health_board = HBName,
         alcohol_admissions = measure)

glimpse(admission_data)
```

## Save dataframe
To save the dataframe to a CSV file, we use the write_csv function.

```{r save-data}
write_csv(admission_data,"alcohol_related_admissions.csv")

```

## Let's compare Grampian and Fife data in 2019

We will use histograms, Q-Q plots and boxplots, before doing a statistical comparison.

```{r compare-Grampian-Fife}

grampian_fife_data <- admission_data %>%                          
  filter(year == 2019,      # only 2019
         health_board %in% c("Grampian", "Fife"))  # only Grampian and Fife

#histogram
grampian_fife_data %>%                         
  ggplot(aes(x = alcohol_admissions)) + 
  geom_histogram(bins = 15) +
  facet_wrap(~health_board)

# Q-Q plot
grampian_fife_data %>%                          
  ggplot(aes(sample = alcohol_admissions)) +          
  geom_qq() + 
  geom_qq_line(colour = "blue") + 
  facet_wrap(~health_board)

# boxplot
grampian_fife_data %>%                          
  ggplot(aes(x = health_board,             
             y = alcohol_admissions)) +    
  geom_boxplot() +                        
  geom_jitter(alpha = 0.4) +               # add data points
  theme(legend.position = "none")          # remove legend


```

## Statistical comparison
Non-parametric test

```{r wilcox-test}
wilcox.test(alcohol_admissions ~ health_board, data = grampian_fife_data)

```
```{r kruskal-test}
# non-parametric test for comparing more than two groups
admission_data %>% 
  filter(year == 2019,
         health_board %in% c("Grampian", "Fife", "Lothian")) %>% 
  kruskal.test(alcohol_admissions~health_board, data = .)

```

## Transformation to normal

```{r log-transform}

Grampian2019 <- admission_data %>%     # save as Grampian2019
  filter(year == 2019,            # only 2019
         health_board == "Grampian") %>%     # only Grampian
  mutate(alcohol_admissions_log = log(alcohol_admissions))         # log hospital admissions

Grampian2019 %>% 
  ggplot(aes(x = alcohol_admissions)) +             
  geom_histogram(bins = 15) +           # make histogram
  ggtitle("Grampian alcohol-related hospital admissions in 2019") # add title

Grampian2019 %>% 
  ggplot(aes(x = alcohol_admissions_log)) +             
  geom_histogram(bins = 15) +           # make histogram
  ggtitle("Log of Grampian alcohol-related hospital admissions in 2019") # add title


```


